\section{Type inference}
\label{sec:type-inference}

In this section we present a type inference algorithm for
mini-Haskell, and discuss soundness and completeness of type inference
with respect to the type system.

For this, we need to consider functional counterparts of relations
used in the type system (Section \ref{Optional-type-classes}). The
main one is satisfiability, the counterpart of entailment (Subsection
\ref{sec:satisfiability}).  We do not consider in this paper
algorithms for constraint set simplification; interested readers may
consult~\cite{OutsideIn2011}. In an abuse of notation the same symbols
are used in this section for functions corresponding to the relations
of constraint set simplification ($\simplifies{P}$, where $P$ is a
program theory) and type generalization (\gen).

We use partial orders on types, constraints, substitutions, and typing
contexts with program theories, defined in Figure \ref{Order}.

Type ordering disregards constraint set entailment, which is important
only for considering whether a constraint $\pi$ can be removed from a
constraint $C$ occurring in a constrained type $C \Rightarrow \tau$;
$\pi$ can be removed if and only if overloading for $\pi$ has been
resolved and there exists a single entailing substitution for $C$, as
defined in Figure \ref{fig:constraint-set-improvement}, page
\pageref{fig:constraint-set-improvement}.

\begin{figure}
   \[ \begin{array}{cc}
   	\displaystyle\frac
          {}
          {\sigma \leq \phi\,\sigma}
   	  & 
   	\displaystyle\frac
   	  {}
   	  {\pi \leq \phi\,\pi} \\[.4cm]
        \multicolumn{2}{c}{
          \displaystyle\frac
	    {\Gamma(x) \leq \Gamma'(x) \text{ for all } x\in \dom{\Gamma}}
	    {P;\Gamma \leq P;\Gamma'}}\\[.4cm]
        \multicolumn{2}{c}{
          \displaystyle\frac
          {\text{there exists $\phi_1$ such that $\phi = \phi_1 \circ \phi'$}}
          {\phi\leq \phi'}}
  \end{array} \]
\caption{Partial orders}
\label{Order}
\end{figure}

A type inference algorithm for core-Haskell is presented in Figure
\ref{Type-inference-fig}, using rules of the form $P;\Gamma \ti
e:(\delta,\phi)$, which means that $\delta$ is the least (principal)
type derivable for $e$ in typing context $\phi\Gamma$ and program
theory $P$, where $\phi\Gamma \leq \Gamma$ and, whenever $\Gamma' \leq
\Gamma$ is such that $P;\Gamma' \ti e: (\delta',\phi')$, we have that
$\phi\Gamma \leq \Gamma'$ and $\delta' \leq \phi\delta$. Furthermore,
we have that $P;\phi\,\Gamma \ti e:(\delta,\phi')$ holds whenever
$P;\Gamma \ti e:(\delta,\phi)$ holds, where $\phi'\leq \phi$
(cf.~theorem \ref{Minimal-type-minimal-typing-context} below).

\begin{Example} {\rm
Consider expression $x$ and ty\-ping context $\Gamma = \{ f: \Int
\rightarrow {\it Int}, x:\alpha \}$; we can derive $\Gamma \ti f\: x:
(\Int, \phi)$, where $\phi = [\alpha \mapsto \Int]$. From
$\phi\,\Gamma = \{ f: \Int \rightarrow \Int, x:\Int \}$, we can derive
$\phi\,\Gamma \ti f\: x:(\Int,\id)$.}
\end{Example}

\begin{Theorem}

If $P;\Gamma \ti e:(\delta,\phi)$ holds then $P;\phi\,\Gamma \ti
e:(\delta,\phi')$ holds, where $\phi'\leq \phi$.

\label{Minimal-type-minimal-typing-context}
\end{Theorem}

\begin{figure}
\[ \begin{array}{cc}
      \displaystyle\frac
        {\begin{array}[t]{ll}
            (\Gamma(\self)(x) = \forall\,\overline{a}.\,\delta) \in \Gamma &
            \overline{b} \mbox{{\rm fresh}}
         \end{array}}
        {P;\Gamma \ti x: (\delta[\overline{a} \mapsto \overline{b}], \id)} (\VVARi) \\ \\

	\displaystyle\frac
          {\begin{array}[t]{lll}
            P;(\Gamma,x:a) \ti e: (C \Rightarrow \tau,\phi) & a\: \mbox{{\rm fresh}} & \tau' = \phi\,a
           \end{array}}
	  {P;\Gamma \ti \lambda x.\:e: (C\Rightarrow \tau' \rightarrow \tau, \phi)} (\ABSi) \\ \\

	\displaystyle\frac
	 {\begin{array}[t]{ll}
             P;\Gamma \ti e: (C \Rightarrow \tau_1, \phi_1) & P;\phi_1\Gamma \ti e': (C' \Rightarrow \tau_2, \phi_2)\\
             \phi' = \mguI(\tau_1 = \tau_2\rightarrow a) & a\: \mbox{{\rm fresh}}, \:
                \phi = \phi' \circ \phi_2 \circ \phi_1\:\: \  \\
             \tau = \phi\:a, \: V = \tv(\tau) \cup \tv(\phi C) & (\phi C \oplus_V\,\phi C') \simplifies{P_\Gamma} D
           \end{array}}
	  {\Gamma \ti e\:e': (D\Rightarrow \tau,\phi)} (\APPi) \\ \\

	\displaystyle\frac
	  {\begin{array}{ll}
             \Gamma \ti e\!:(C\Rightarrow \tau,\phi_1) & C \simplifies {P_\Gamma} C' \\
             \gen(\sigma,C'\Rightarrow \tau,\tv(\phi_1\Gamma)) & \phi_1\Gamma,\,x\!:\!\sigma \ti e_2\!:(\delta,\phi)
          \end{array}}
	 {\Gamma \ti \mlet\ x=e\ \iin\ e':(\delta,\phi)} (\LETi)
\end{array} \]
\caption{Type Inference}
\label{Type-inference-fig}
\end{figure}

Relation $\mgu$ is the most general (least) unifier
relation~\cite{Robinson65}: $\mgu(\Tau,\phi)$ is defined to hold
between a set of pairs of simple types $\Tau$ and a substitution
$\phi$ if i) $\phi$ is a unifier of every pair in $\Tau$ (i.e.~$\phi
\tau = \phi\tau'$ for every $(\tau,\tau')\in \Tau$), and ii) it is the
least such unifier (i.e.~if $\phi'$ is a unifier of all pairs in
$\Tau$, then $\phi\leq \phi'$). The relation holds similarly for
constraints instead of types.

$\mguI$ is a function that gives a most general unifier of a set of
pairs of simple types (or simple constraints). We define also that
$\phi = \mguI(\tau = \tau')$ is an alternative notation for
$\phi = \mguI(\{(\tau, \tau')\})$. We have:

\begin{Theorem}[Soundness]

If $P;\Gamma \ti e: (\delta,\phi)$ holds then $P;\phi\Gamma \vdash e: \delta$ holds.

\label{thm:type-inference-sound}
\end{Theorem}

\begin{Theorem}[Principal type]

If $P;\Gamma \ti e: (\delta,\phi)$ holds then, for all $\delta'$ such
that $P;\phi\Gamma \vdash e: \delta'$ holds, we have that $\delta \leq
\delta'$.

\label{thm:principal-type}
\end{Theorem}

A completeness theorem does not hold. For example, the canonical
Haskell ambiguity example of expression $e_0 = (\sshow\ $\!\!$.$\!\!$
\rread))$ --- where \sshow\ has type \SShow a $\Rightarrow$ $a
\rightarrow$ \String, and \rread\ has type \RRead a $\Rightarrow$
\String $\rightarrow a$ ---, we have that there exists $P$ and
$\Gamma$ such that $P;\Gamma \vdash e_0: \String \rightarrow \String$
holds, but there is no $\delta,\phi$ such that $P;\Gamma \vdash e_0:
(\delta,\phi)$ holds.

The greater simplicity obtained by allowing type instantiation to
occurr in a context-independent way, in a type system for a language
with support for context-dependent overloading, has significant
counterparts. The disadvantages are: ambiguous expressions are allowed
to be well-typed, and there exist several translations for
expressions, one of them a principal translation, for a semantics
defined inductively on the type system rules.

A declarative specification of type inference, with a unique type
derivable for each expression, where type instantiation is restricted
to be done only in a context-dependent way, defined by considering
functions used in the type inference algorithm as relations, is a
possible alternative. In this case, the type inference algorithm is
obtained directly from a declarative specification of the type system
by transforming relations used into functions. The fact that every
element has a unique type is consonant with everyday spoken
language. It is straightforward to define, a posteriori, the set of
types that are valid instances of the type of an expression.

The fact that only a single type can be derived for each expression
rules out the possibility of having distinct derivations for an
expression's type. Thus, an error message for an expression such as
{\tt (\sshow\ $\!\!$.$\!\!$ \rread)}, in a context with more than one
instance for \SShow\ and \RRead, should be that the expression can not
be given a well-defined semantics (there is no type that would allow
it to have a well-defined semantics). Distinct meanings of {\tt
  (\sshow\ $\!\!$.$\!\!$ \rread)} would be obtained from distinct
instance types of \sshow\ and \rread.

Type inference for mini-Haskell is obtained by extending type
inference for core-Haskell in straightforward way, namely by directly
using the rules of Subsection \ref{sec:mini-Haskell} (Figures
\ref{fig:mini-haskell-module-rule}, \ref{fig:import-relation} and
\ref{fig:mini-haskell-rules-for-declarations}), by replacing relations
with functions.

The overloading resolution theorem below considers a type inference
algorithm that differs from mini-Haskell's by only i) disregarding
improvement, and thus not removing any {\em resolved\/} constraint,
and ii) not allowing constraints in an argument to be removed, using
$C \cup C'$ instead of $C \oplus_V C'$ (where $V = \tv(\tau)\cup
\tv(C)$). We use $\tiUm$ instead of $\ti$ in typing formulas of this
(mini-Haskell without improvement and constraint selection by
$\oplus$) type inference algorithm. We also define a program context
$\CO[e]$ as any expression that has $e$ as a subexpression.

\begin{Theorem}[Overloading Resolution]
  For all $P, \Gamma, e, C, \tau$ such that $P;\Gamma \tiUm e:
  C\Rightarrow \tau$ holds, $\pi\in C$ and
  $a\in\unreachableVars(\{\pi\},\tv(\phi(\tau)))$, then, for all
  $P;\phi\Gamma\tiUm\CO[e]: C'\Rightarrow \tau'$ that holds, we have
  that $\pi \in C'$.
\label{thm:overloading-resolution}
\end{Theorem}

{\em Proof\/}: By induction on the structure of $\CO[e]$. $\qed$

Informally speaking, theorem \ref{thm:overloading-resolution} shows
that there is no program context where an expression can be used that
will cause a constraint with an unreachable type variable to be
instantiated.

\input{satisfiability}
